\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-11-24 AbstractPeerJ.tex
abstract edits
edits to abstract for peerj

The \emph{teosinte branched1} (\emph{tb1}) gene is a major QTL controlling branching differences between maize and its wild progenitor, teosinte. The insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} is known to enhance the gene's expression, causing reduced tillering in maize. Observations of the maize \emph{tb1} allele in teosinte and estimates of an insertion age of the \emph{Hopscotch} that predates domestication led us to investigate its prevalence and potential role in teosinte. We assessed the prevalence of the \emph{Hopscotch} element across an Americas-wide sample of 837 maize and teosinte individuals using a co-dominant PCR assay. Additionally we calculated population genetic summaries using sequence data from a subset of individuals from four teosinte populations and collected phenotypic data using seed from a single teosinte population where \emph{Hopscotch} was found segregating at high frequency. Genotyping results indicate the \emph{Hopscotch} element is found in a number of teosinte populations and linkage disequilibrium near \emph{tb1} does not support recent introgression from maize. Population genetic signatures are consistent with selection on this locus revealing a potential ecological role for \emph{Hopscotch} in teosinte, but a greenhouse experiment does not detect a strong association between \emph{tb1} and tillering in teosinte.

\emph{Conclusions:} Our findings suggest the role of \emph{Hopscotch} differs between maize and teosinte. Future work should assess \emph{tb1} expression levels in teosinte with and without the \emph{Hopscotch} and more comprehensively phenotype teosinte to assess the ecological significance of the \emph{Hopscotch} insertion and, more broadly, the \emph{tb1} locus in teosinte.

\vspace*{0.2cm}

>>>>

The \emph{teosinte branched1} (\emph{tb1}) gene is a major QTL controlling branching differences between maize and its wild progenitor, teosinte. The insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} is known to enhance the gene's expression, causing reduced tillering in maize. Observations of the maize \emph{tb1} allele in teosinte and estimates of an insertion age of the \emph{Hopscotch} that predates domestication led us to investigate its prevalence and potential role in teosinte. We assessed the prevalence of the \emph{Hopscotch} element across an Americas-wide sample of 837 maize and teosinte individuals using a co-dominant PCR assay. Additionally we calculated population genetic summaries using sequence data from a subset of individuals from four teosinte populations and collected phenotypic data using seed from a single teosinte population where \emph{Hopscotch} was found segregating at high frequency. Genotyping results indicate the \emph{Hopscotch} element is found in a number of teosinte populations and linkage disequilibrium near \emph{tb1} does not support recent introgression from maize. Population genetic signatures are consistent with selection on this locus revealing a potential ecological role for \emph{Hopscotch} in teosinte, but a greenhouse experiment does not detect a strong association between \emph{tb1} and tillering in teosinte.
Our findings suggest the role of \emph{Hopscotch} differs between maize and teosinte. Future work should assess \emph{tb1} expression levels in teosinte with and without the \emph{Hopscotch} and more comprehensively phenotype teosinte to assess the ecological significance of the \emph{Hopscotch} insertion and, more broadly, the \emph{tb1} locus in teosinte.\\
